Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-07-11 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

The MR expression (Eqn.~\ref{eq:MR}) contains phase saturations, $S_{w}\left(x_{i},t\right)$ and  $S_{nw}\left(x_{i},t\right)$ (Eqns.~\ref{Eqn:CoreyBrooks1}-\ref{Eqn:CoreyBrooks2}), however it is clear that during fluid displacement they change in time and space according to the saturation equation (Eqn.~\ref{eqn:saturation_eqn}). Assuming that phase viscosities remain invariant throughout the simulated fluid displacement, in this work, with no lack of generality, the MR can be replaced by the viscosity ratio (VR),
\begin{displaymath} 
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
that will be used in the parametrisation of the numerical simulations conducted in Section \ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent prognostic field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$ variables.

\medskip
In the next section, the numerical formulation used to simulate multi-fluid flow in porous media is briefly validated (Section \ref{section:results_initial_model_validation}) against laboratory experiments (qualitative validation). The impact of VR (quantitative validation) and heterogeneity on the onset instability and growth of fingers are numerically investigated in Section \ref{section:results_homo_hete}. Section \ref{section:results_hete_fix_adapt} demonstrates the importance of an appropriate mesh resolution to adequately capture the initial stages of viscous fingers formation and development. Finally, flow pathway (channelling) is the focus of Section \ref{section:results_3D}.

@@

%%%%%%%%%%%% 
\subsection{Model Set-up}\label{section:results_setup} 
Numerical simulations were conducted with the model summarised in Section \ref{equations_scheme} and embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} \citep[a full description of the model can be found in][]{jackson_2013,gomes_2017}. This multi-physics model has been validated against traditional multi-fluids test-cases (\eg advection-diffusion, Buckley-Leverett problem, channel model, immiscible displacement, gravity-driven displacement etc) in \citet{radunz_2014}, \citet{jackson_2015}, \citet{salinas2015} and \citet{pavlidis2016}.

\medskip
In this work, all test-cases were performed in idealised geometries discretised with unstructured triangular and tetrahedral mesh using the \PN[1]{2}, \PN[1]{1} (Fig.~\ref{fig:fem_elem}, Sections~\ref{section:results_initial_model_validation}-\ref{section:results_hete_fix_adapt}) and \PNDG[1]{1} (Section~\ref{section:results_3D})  FE-pairs. An implicit Crank-Nicolson time-stepping scheme was used with {\it a posteriori} adaptive time-step size targeting a maximum Courant-Friedrichs-Lewy condition \citep[CFL,][]{courant_1941} of 2. In most simulations, the domain was initially fully saturated with non-wetting fluid which was displaced by a pure (wetting) fluid at a prescribed initial velocity $\left(u^{\circ}\right)$. For simplicity, the porosity ($\phi$) of the domain was kept constant at 0.2 in all simulations, whereas the absolute permeability ($\mathbf{K}$) varied in space, \ie $\mathbf{K}=\mathbf{K}\left(x_{i}\right)$. Fluids are assumed incompressible, also gravity and capillary pressure were neglected. Initial set-up for the numerical simulations is summarised in Table \ref{table:setup}.

@@

\medskip
%A numerical simulation was also performed in modified conditions (see Fig.~\ref{fem_cv_represent_a}) to assess model behaviour under larger {\it VR}=10. Initial and boundary conditions remain the same, and flow behaviour qualitatively agree validated against laboratory experiments conducted by \citet{evans_1994}. 
A two-phase immiscible flow along homogeneous and parallel layers of contrasting petrophysical properties (\ie permeability) is initially simulated. During the fluid displacement, crossflow between adjacent layers often occur due to viscous, capillary and/or gravitational forces that drive the flow. In this numerical simulation, crossflow is caused by viscous forces, which is commonly named viscous crossflow. Figures~\ref{fem_cv_represent_a}(b) and (c) show the continuous displacement of fluid 2 due to the injection of fluid 1. They also demonstrate the preferential flow path through high-permeability regions. Such flow behaviour, represented by the crossflow through the four regions, is in good qualitative agreement with experiments conducted by \citet{dawe_2008}.

>>>>

The MR expression (Eqn.~\ref{eq:MR}) contains phase saturations, $S_{w}\left(x_{i},t\right)$ and  $S_{nw}\left(x_{i},t\right)$ (Eqns.~\ref{Eqn:CoreyBrooks1}-\ref{Eqn:CoreyBrooks2}), however it is clear that during fluid displacement they change in time and space according to the global mass balance equation. Assuming that phase viscosities remain invariant throughout the simulated fluid displacement, in this work, with no lack of generality, the MR can be replaced by the viscosity ratio (VR),
\begin{displaymath} 
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
which will be used in the parametrisation of the numerical simulations conducted in Section~\ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent prognostic field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$ variables.

\medskip
In the next section, the numerical formulation used to simulate multi-fluid flow in porous media is briefly validated (Section \ref{section:results_initial_model_validation}) against laboratory experiments (qualitative validation). The impact of VR (quantitative validation) and heterogeneity on the onset instability and growth of fingers are numerically investigated in Section \ref{section:results_homo_hete}. Section \ref{section:results_hete_fix_adapt} demonstrates the importance of appropriate mesh resolution to adequately capture the initial stages of viscous fingers formation and development. Finally, flow pathway (channelling) is the focus of Section \ref{section:results_3D}.

@@

%%%%%%%%%%%% 
\subsection{Model Set-up}\label{section:results_setup} 
Numerical simulations were conducted with the model summarised in Section \ref{equations_scheme} and embedded in the next-generation flow simulator Fluidity/IC-FERST model software\footnote{\href{http://multifluids.github.io}{http://multifluids.github.io}} \citep[a full description of the model can be found in][]{gomes_2017,salinas_2018}. This multi-physics model has been validated against traditional multi-fluids test-cases (\eg advection-diffusion, Buckley-Leverett problem, channel model, immiscible displacement, gravity-driven displacement etc) in \citet{radunz_2014}, \citet{jackson_2015}, \citet{salinas2015} and \citet{pavlidis2016}.

In this work, all test-cases were performed in idealised geometries discretised with unstructured triangular and tetrahedral mesh using the \PN[1]{2}, \PN[1]{1} (Fig.~\ref{fig:fem_elem}, Sections~\ref{section:results_initial_model_validation}-\ref{section:results_hete_fix_adapt}) and \PNDG[1]{1} (Section~\ref{section:results_3D})  FE-pairs. An implicit Crank-Nicolson time-stepping scheme was used with {\it a posteriori} adaptive time-step size targeting a maximum Courant-Friedrichs-Lewy condition \citep[CFL,][]{courant_1941} of 2. In most simulations, the domain was initially fully saturated with non-wetting fluid which was displaced by a pure (wetting) fluid at a prescribed initial velocity $\left(u^{\circ}\right)$. For simplicity, the porosity ($\phi$) of the domain was kept constant at 0.2 in all simulations, whereas the absolute permeability ($\mathbf{K}$) varied in space, \ie $\mathbf{K}=\mathbf{K}\left(x_{i}\right)$. Fluids are assumed incompressible, also gravity and capillary pressure were neglected. Initial set-up for the numerical simulations is summarised in Table \ref{table:setup}.

@@

%A numerical simulation was also performed in modified conditions (see Fig.~\ref{fem_cv_represent_a}) to assess model behaviour under larger {\it VR}=10. Initial and boundary conditions remain the same, and flow behaviour qualitatively agree validated against laboratory experiments conducted by \citet{evans_1994}. 
A two-phase immiscible flow along homogeneous and parallel layers of contrasting petrophysical properties (\ie permeability) is initially simulated. During fluid displacement, crossflow between adjacent layers often occur due to viscous, capillary and/or gravitational forces that drive the flow. In this numerical simulation, crossflow is caused by viscous forces, which is commonly named viscous crossflow. Figures~\ref{fem_cv_represent_a}(b) and (c) show continuous displacement of fluid 2 due to the injection of fluid 1. They also demonstrate the preferential flow path through high-permeability regions. Such flow behaviour, represented by the crossflow through the four regions, is in good qualitative agreement with experiments conducted by \citet{dawe_2008}.
